---
title: "Carpet Methods Code"
author: 
output: 
  html_document:
  smart: no
theme: flatly
float: yes
css: "style.css"    
editor_options: 
  chunk_output_type: console
mainfont: Arial
mathfont: Arial
fontsize: 12pt
subtitle:
---

# {.tabset .tabset-pills .tabset-fade}
  
## Prolog


```{r, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}

# PROLOG   ##############################################################################

# PROJECT: Seed Rain Methods
# PURPOSE: Data analysis - Dryad
#
# AUTHORS: Lauren Sullivan (llsull@msu.edu) & Lauren Shoemaker (lshoema1@uwyo.edu)
# COLLAB:  Melissa DeSiervo, Larissa Kahan
# EDITED:  19 Jan 2026
#
# FILES:   1) seed_recovery_all.csv
#
# NOTES:   runs statistical analysis, plots exploratory figures and model-based figures
#
# NEXT STEPS: estimate slopes from mass by recovery fig (Fig 3), decide on MO post-sticky


# PROLOG   ##############################################################################
```
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# clear workspace
rm(list=ls())

# working directory
# set working directory to "source file location"

# libraries
library(tidyverse)
library(RColorBrewer)
library(viridis)
library(cowplot)
library(glmmTMB)
library(ggeffects) #for ggpredict()
library(emmeans) # for post-sticky comparison

# data
all <- read_csv("seed_recovery_all.csv")

```

## Exploratory Figures

Exploratory figures of recovery rates by collection type and by site.

First lets look at the percent recovered.

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}


palatte <- c("#CC79A7","#E69F00")

#size order
mo_raw <- subset(all,site == "mo" & treatment != "post-sticky")
mo_raw$species <- factor(mo_raw$species, levels = c("TRIPER", "CORTIN", "RUDHIR", "SCHSCO", "SPOHET", "LESCAP", "DESCAN", "ECHPAL"), labels = c("T. perfoliata", "C. tinctoria", "R. hirta", "S. scoparium", "S. heterolepis", "L. capitata", "D. canadense", "E. pallida"))
mo_raw$treatment <- factor(mo_raw$treatment, levels = c("carpet", "sticky"), labels = c("artificial\ngrass", "sticky"))

mo_raw <- ggplot(mo_raw, aes(x = as.factor(time_month), y = pct_recovered, fill = treatment))+
  geom_boxplot(alpha = 0.5)+
  facet_grid(cols = vars(species))+
  scale_fill_manual(values = palatte, name = "Trap Type")+
  theme_bw()+
  labs(y = "Percent Seeds Recovered")+
  theme(strip.text.x = element_text(size = 10, face = "italic"),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14, vjust = +1.5),
        axis.text.y = element_text(size = 11),
        axis.text.x = element_text(size = 11)
      )


#size order
co_raw <- subset(all, site == "co")
co_raw$species <- factor(co_raw$species, levels = c("b.gracilis", "s.rigida", "s.angustifolium", "d.purpurea", "h.annuus", "a.fruticosa", "b.dactyloides", "y.glauca"), labels = c("B. gracilis", "S. rigida", "S. angustifolium", "D. purpurea", "H. annuus", "A. fruticosa", "B. dactyloides", "Y. glauca"))
co_raw$treatment <- factor(co_raw$treatment, levels = c("carpet", "sticky"), labels = c("artificial\ngrass", "sticky"))

co_raw <- ggplot(co_raw, aes(x = as.factor(time_month), y = pct_recovered, fill = treatment))+
  geom_boxplot(alpha = 0.5)+
  facet_grid(cols = vars(species))+
  scale_fill_manual(values = palatte, name = "Trap Type")+
  theme_bw()+
  labs(y = "Percent Seeds Recovered",x = "Time (month)")+
  theme(strip.text.x = element_text(size = 10, face = "italic"),
        axis.title.y = element_text(size = 14, vjust = +1.5),
        axis.title.x = element_text(size = 14, vjust = -0.75),
        axis.text.y = element_text(size = 11),
        axis.text.x = element_text(size = 11)
      )

plot_grid(mo_raw, co_raw, ncol = 1, labels = c("A)", "B)"))

## save figure
pdf("../figures/all_raw_data.pdf", width = 11, height = 8)

plot_grid(mo_raw, co_raw, ncol = 1, labels = c("A)", "B)"), align = "h", axis = "l")

invisible(dev.off())

```

Next let's look at the LRR of recovery of carpets relative to sticky traps, so positive values mean the sticky traps retain more, and the negative values mean the carpets retain more.



```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

# make relativized capture plot with carpet compared to sticky.

dat_small <- all[,1:8]
dat_small <- subset(dat_small, treatment != "post-sticky")

dat_wide <- dat_small %>%
  pivot_wider(
    names_from = treatment,
    values_from = number_recovered
  )

dat_wide$LRR_removed <- log(dat_wide$sticky / dat_wide$carpet)

#size order
mo_wide <- subset(dat_wide, site == "mo")
mo_wide$species <- factor(mo_wide$species, levels = c("TRIPER", "CORTIN", "RUDHIR", "SCHSCO", "SPOHET", "LESCAP", "DESCAN", "ECHPAL"), labels = c("T. perfoliata", "C. tinctoria", "R. hirta", "S. scoparium", "S. heterolepis", "L. capitata", "D. canadense", "E. pallida"))


mo_rel <- ggplot(mo_wide, aes(x = as.factor(time_month), y = LRR_removed))+
  geom_boxplot()+
  geom_hline(yintercept = 0, linetype = "dashed", color = "red")+
  facet_grid(cols = vars(species))+
  scale_color_brewer(palette = "Set1")+
  theme_bw()+
  ylab("Recovery Log Response Ratio")+
  theme(strip.text.x = element_text(size = 10, face = "italic"),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14, vjust = +1.5),
        axis.text.y = element_text(size = 11),
        axis.text.x = element_text(size = 11)
      )

#size order
co_wide <- subset(dat_wide, site == "co")
co_wide$species <- factor(co_wide$species, levels = c("b.gracilis", "s.rigida", "s.angustifolium", "d.purpurea", "h.annuus", "a.fruticosa", "b.dactyloides", "y.glauca"), labels = c("B. gracilis", "S. rigida", "S. angustifolium", "D. purpurea", "H. annuus", "A. fruticosa", "B. dactyloides", "Y. glauca"))

co_rel <- ggplot(co_wide, aes(x = as.factor(time_month), y = LRR_removed))+
  geom_boxplot()+
  geom_hline(yintercept = 0, linetype = "dashed", color = "red")+
  facet_grid(cols = vars(species))+
  scale_color_brewer(palette = "Set1")+
  theme_bw()+
  ylab("Recovery Log Response Ratio")+
  xlab("Time (month)")+
  theme(strip.text.x = element_text(size = 10, face = "italic"),
        axis.title.y = element_text(size = 14, vjust = +1.5),
        axis.title.x = element_text(size = 14, vjust = -0.75),
        axis.text.y = element_text(size = 11),
        axis.text.x = element_text(size = 11)
      )

plot_grid(mo_rel, co_rel, ncol = 1, labels = c("A)", "B)"))


## save figure
pdf("../figures/all_relative_data.pdf", width = 10, height = 8)

plot_grid(mo_rel, co_rel, ncol = 1, labels = c("A)", "B)"), align = "h", axis = "l")

invisible(dev.off())

```

## Analysis

Explore how proportion seed removal is a function of seed size, trap method and time with a random effect for block nested within site.

```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
dat_all <- subset(all, treatment != "post-sticky")

# Statistical Model
mod1 <- glmmTMB(pct_recovered ~ treatment * mean_1seed + time_month + (1|site/block), 
                data = dat_all, 
                weights = number_original, 
                family = binomial)
summary(mod1) 


```



Summary stats and main text figure to represent the significant trap type * seed mass interaction using model predicted values.

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

#Predict the data from the statistical model
out.pred <- predict(mod1, type = 'response')


predicted <- data.frame(pred = out.pred,
                        mass = dat_all$mean_1seed,
                        time = dat_all$time_month,
                        treatment = dat_all$treatment,
                        num_rec = dat_all$number_recovered,
                        num_lost = dat_all$number_lost)

## summary stats
predicted %>%
  group_by(treatment) %>%
  dplyr::summarize(mean_recovered = mean(pred),
                   sd_recovered = sd(pred))


palatte <- c("#CC79A7","#E69F00" )

predicted$treatment <- factor(predicted$treatment, levels = c("carpet", "sticky"), labels = c("artificial grass", "sticky"))



#graph predicted data and binomial smooths
mass <- ggplot(predicted, aes(x = mass, y = pred, 
                     color = treatment))+
          geom_point(position = position_jitter(width = 0.0002), alpha = 0.5)+
          theme_bw()+
          ylim(0,1)+
          scale_colour_manual(values = palatte)+
          labs(y= "Proportion of Seeds Recovered", x = "Seed mass", color = "Trap Type")+ 
          stat_smooth(method = "glm",
                      method.args = list(family = "binomial"))+  
          theme(axis.title.y = element_text(size = 14, vjust = +1.5),
                axis.title.x = element_text(size = 14, vjust = -0.75),
                axis.text.y = element_text(size = 11),
                axis.text.x = element_text(size = 11)
      )

mass

## save figure
pdf("../figures/recovery_by_mass.pdf", width = 6, height = 5)

mass

invisible(dev.off())

```


Examining model output and different comparisons we want to report and visualize. In particular, use emmeans to examine comparisons among trap types at different seed masses.

```{r}

# Contrasts for different seed masses
EM <- emmeans::emmeans(mod1, ~treatment |mean_1seed, at = list(mean_1seed = c(0)))
pairs(EM) ### average results over random effects ##

EM2 <- emmeans::emmeans(mod1, ~treatment |mean_1seed, at = list(mean_1seed = c(0.02)))
pairs(EM2) ### average results over random effects ##

```


Post-sticky analysis from MO.  How seed recovery varies across trap type when traps are out for one month (stick, artificial grass and post-sticky)

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
dat_ps <- all[,1:8]
dat_ps <- subset(dat_ps, site == "mo" & time_month == 1)

dat_ps$pct_recovered <- dat_ps$number_recovered/dat_ps$number_original



dat_ps$treatment <- factor(dat_ps$treatment, levels = c("post-sticky", "carpet", "sticky"), labels = c("post-sticky", "artificial grass", "sticky"))
```

```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
mod2 <- glmmTMB(pct_recovered ~ treatment + (1|block) + (1|species), 
                data = dat_ps, 
                family = binomial)
summary(mod2) 
```

Figure with predicted data

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
#Predict the data from the statistical model
out.pred2 <- predict(mod2, type = 'response')


predicted2 <- data.frame(pred2 = out.pred2,
                        treatment = dat_ps$treatment)


stats2 <- predicted2 %>%
            group_by(treatment) %>%
            dplyr::summarize(mean_pct = mean(pred2),
                             sd_pct = sd(pred2))

post_sticky <- ggplot(stats2, aes(x=treatment, y = mean_pct))+
          geom_point(size=4)+
          geom_errorbar(aes(ymin = mean_pct - sd_pct, ymax = mean_pct + sd_pct), width = 0.4, linewidth = 1)+
          theme_bw()+
          ylim(0,1)+
          labs(y= "Proportion of Seeds Recovered", x = "Trap Type")+ 
          theme(axis.title.y = element_text(size = 14, vjust = +1.5),
                axis.title.x = element_text(size = 14, vjust = -0.75),
                axis.text.y = element_text(size = 11),
                axis.text.x = element_text(size = 11)
      )

post_sticky


## save figure
pdf("../figures/post_sticky.pdf", width = 5, height = 5)

post_sticky

invisible(dev.off())

```


## Supplemental figure 

Explore the predicted outcome for all factors (trap type, mass, time and site)

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

dat_full_renamed <- dat_all %>%
  mutate(Month = time_month) %>%
  select(-time_month)
mod2 <- glmmTMB(pct_recovered ~ treatment * mean_1seed + Month + (1|site/block), 
                data = dat_full_renamed, 
                weights = number_original, 
                family = binomial)
summary(mod2) 
# Predicted figure for all factors
pred <- ggpredict(mod2, terms=c("mean_1seed", "treatment", "site", "Month"), type="random")

pred$facet <- ifelse(pred$facet == "mo", "Missouri", "Colorado")
pred_all <- plot(pred) +
  labs(x="Seed mass (g)", y="Percent Recovered", color = "Trap Type") +
  ggtitle("") +
  scale_color_manual(values = c("#CC79A7","#E69F00" ), labels = c("artificial grass", "sticky")) +
  scale_fill_manual(values = c("#CC79A7","#E69F00" )) +  
          theme(axis.title.y = element_text(size = 14, vjust = +1.5),
                axis.title.x = element_text(size = 11, vjust = -0.75),
                axis.text.y = element_text(size = 11),
                axis.text.x = element_text(size = 11, angle=45, vjust=.7))
      
pred_all


## save figure
pdf("../figures/all_predicted.pdf", width = 5, height = 6)

pred_all

invisible(dev.off())

```
