---
title: "Carpet Methods Code"
author: 
output: 
  html_document:
  smart: no
theme: flatly
float: yes
css: "style.css"    
editor_options: 
  chunk_output_type: console
mainfont: Arial
mathfont: Arial
fontsize: 12pt
subtitle:
---

# {.tabset .tabset-pills .tabset-fade}
  
## Prolog


```{r, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}

# PROLOG   ##############################################################################

# PROJECT: Seed Rain Methods
# PURPOSE: Data cleaning 
#
# AUTHORS: Lauren Sullivan (llsull@msu.edu) & Lauren Shoemaker (lshoema1@uwyo.edu)
# COLLAB:  Melissa DeSiervo, Larissa Kahan
# EDITED:  20 April 2025
#
# FILES:   1) data/L0/MO_seedrain_raw.csv
#          2) data/LO/MO_seedweight_raw.csv
#          3) data/L0/CO_seedrain_raw.csv
#          4) data/L0/CO_seedweight_raw.csv
#
# NOTES:   creates "strict" datasets where we just focus on the species of interest 
#             not herbivory or other species that were collected in the traps,
#             runs statistical analysis, plots exploratory figures and model-based figures
#
# NEXT STEPS: estimate slopes from mass by recovery fig (Fig 3), decide on MO post-sticky


# PROLOG   ##############################################################################
```
```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# clear workspace
rm(list=ls())

# working directory
# set working directory to "source file location"

# libraries
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(glmmTMB)
library(ggeffects) #for ggpredict()
library(emmeans) # for post-sticky comparison

# data
mo <- read_csv("../data/L0/MO_seedrain_raw.csv")
mo_mass <- read_csv("../data/L0/MO_seedweight_raw.csv")

co <- read_csv("../data/L0/CO_seedrain_raw.csv")
co_mass <- read_csv("../data/L0/CO_seedweight_raw.csv")

```
## Data Cleaning

Clean and combine all data from CO and MO sites - both the recovery and the seed mass data.

Missouri data cleaning

```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}

# clean up treatment names and times

mo$time_month[mo$time_month == '1 month'] <- 1
mo$time_month[mo$time_month == '2 month'] <- 2
mo$time_month[mo$time_month == '1 week'] <- 0.25

mo$time_month <- as.numeric(mo$time_month)


# clean recovery data to just be the species we selected, not the extra 
#      seeds we caught throughout the experiment

mo_slim <- mo %>%
  filter(species == "LESCAP" | species == "DESCAN" | species == "TRIPER" |
         species == "SCHSCO" | species == "ECHANG" | species == "SPOHET" |
         species == "RUDHIR" | species == "CORTIN")


#we used ECHPAL, not ECHANG - so update data
mo_slim$species[mo_slim$species == 'ECHANG'] <- "ECHPAL"

# clean seed mass data

mo_mass_small <- mo_mass %>% 
  group_by(species) %>%
  dplyr::summarize(mean_1seed = mean(lot_mass_g)/25)

# rename species to match recovery data
mo_mass_small$species[mo_mass_small$species == 'RATPIN'] <- "RUDHIR"
mo_mass_small$species[mo_mass_small$species == 'ECHANG'] <- "ECHPAL"

```

Colorado data cleaning

```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}

# clean up treatment names and times

co$treatment[co$treatment == 's'] <- "sticky"
co$treatment[co$treatment == 'c'] <- "carpet"


# clean recovery data to just be the species we selected, not the extra 
#      seeds we caught throughout the experiment

co_slim <- co %>%
  filter(species != ("various"))

co_mass_small <- co_mass %>%
  pivot_longer(
    cols = afrut:yglau,
    names_to = "species",
    values_to = "mass"
  ) %>%
  group_by(species) %>%
  dplyr::summarize(mean_1seed = mean(mass)/25)


#rename species to match recovery data
co_mass_small$species[co_mass_small$species == 'afrut'] <- "a.fruticosa"
co_mass_small$species[co_mass_small$species == 'bdact'] <- "b.dactyloides"
co_mass_small$species[co_mass_small$species == 'bgrac'] <- "b.gracilis"
co_mass_small$species[co_mass_small$species == 'dpurp'] <- "d.purpurea"
co_mass_small$species[co_mass_small$species == 'hannu'] <- "h.annuus"
co_mass_small$species[co_mass_small$species == 'origi'] <- "s.rigida"
co_mass_small$species[co_mass_small$species == 'sangu'] <- "s.angustifolium"
co_mass_small$species[co_mass_small$species == 'yglau'] <- "y.glauca"

```

Join data together


```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
### Create full dataset with both sites

dat_all <- full_join(mo_slim, co_slim)

list(unique(dat_all$species))

# create a %consumed column

dat_all$pct_recovered <- dat_all$number_recovered/dat_all$number_original
dat_all$time_month <- as.factor(dat_all$time_month)


```

## Exploratory Figures

Exploratory figures of recovery rates by collection type and by site.

First lets look at the percent recovered.

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

#size order
mo_raw <- subset(dat_all,site == "mo" & treatment != "post-sticky")
mo_raw$species <- factor(mo_raw$species, levels = c("TRIPER", "CORTIN", "RUDHIR", "SCHSCO", "SPOHET", "LESCAP", "DESCAN", "ECHPAL"), labels = c("T. perfoliata", "C. tinctoria", "R. hirta", "S. scoparium", "S. heterolepis", "L. capitata", "D. canadense", "E. pallida"))


mo_raw <- ggplot(mo_raw, aes(x = time_month, y = pct_recovered, color = treatment))+
  geom_boxplot()+
  facet_grid(cols = vars(species))+
  scale_color_brewer(palette = "Set1")+
  theme_bw()+
  ylab("Percent Seeds Recovered")+
  theme(strip.text.x = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14, vjust = +1.5),
        axis.text.y = element_text(size = 11),
        axis.text.x = element_text(size = 11)
      )


#size order
co_raw <- subset(dat_all, site == "co")
co_raw$species <- factor(co_raw$species, levels = c("b.gracilis", "s.rigida", "s.angustifolium", "d.purpurea", "h.annuus", "a.fruticosa", "b.dactyloides", "y.glauca"), labels = c("B. gracilis", "S. rigida", "S. angustifolium", "D. purpurea", "H. annuus", "A. fruticosa", "B. dactyloides", "Y. glauca"))

co_raw <- ggplot(co_raw, aes(x = time_month, y = pct_recovered, color = treatment))+
  geom_boxplot()+
  facet_grid(cols = vars(species))+
  scale_color_brewer(palette = "Set1")+
  theme_bw()+
  ylab("Percent Seeds Recovered")+
  xlab("Time (month)")+
  theme(strip.text.x = element_text(size = 10, face = "italic"),
        axis.title.y = element_text(size = 14, vjust = +1.5),
        axis.title.x = element_text(size = 14, vjust = -0.75),
        axis.text.y = element_text(size = 11),
        axis.text.x = element_text(size = 11)
      )

plot_grid(mo_raw, co_raw, ncol = 1, labels = c("A)", "B)"))

## save figure
pdf("../figures/all_raw_data.pdf", width = 10, height = 8)

plot_grid(mo_raw, co_raw, ncol = 1, labels = c("A)", "B)"), align = "h", axis = "l")

invisible(dev.off())

```

Next let's look at the LRR of recovery of carpets relative to sticky traps, so positive values mean the sticky traps retain more, and the negative values mean the carpets retain more.



```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

# make relativized capture plot with carpet compared to sticky.

dat_small <- dat_all[,1:8]
dat_small <- subset(dat_small, treatment != "post-sticky")

dat_wide <- dat_small %>%
  pivot_wider(
    names_from = treatment,
    values_from = number_recovered
  )

dat_wide$LRR_removed <- log(dat_wide$sticky / dat_wide$carpet)

#size order
mo_wide <- subset(dat_wide, site == "mo")
mo_wide$species <- factor(mo_wide$species, levels = c("TRIPER", "CORTIN", "RUDHIR", "SCHSCO", "SPOHET", "LESCAP", "DESCAN", "ECHPAL"), labels = c("T. perfoliata", "C. tinctoria", "R. hirta", "S. scoparium", "S. heterolepis", "L. capitata", "D. canadense", "E. pallida"))


mo_rel <- ggplot(mo_wide, aes(x = time_month, y = LRR_removed))+
  geom_boxplot()+
  geom_hline(yintercept = 0, linetype = "dashed", color = "red")+
  facet_grid(cols = vars(species))+
  scale_color_brewer(palette = "Set1")+
  theme_bw()+
  ylab("Recovery Log Response Ratio")+
  theme(strip.text.x = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14, vjust = +1.5),
        axis.text.y = element_text(size = 11),
        axis.text.x = element_text(size = 11)
      )


#size order
co_wide <- subset(dat_wide, site == "co")
co_wide$species <- factor(co_wide$species, levels = c("b.gracilis", "s.rigida", "s.angustifolium", "d.purpurea", "h.annuus", "a.fruticosa", "b.dactyloides", "y.glauca"), labels = c("B. gracilis", "S. rigida", "S. angustifolium", "D. purpurea", "H. annuus", "A. fruticosa", "B. dactyloides", "Y. glauca"))

co_rel <- ggplot(co_wide, aes(x = time_month, y = LRR_removed))+
  geom_boxplot()+
  geom_hline(yintercept = 0, linetype = "dashed", color = "red")+
  facet_grid(cols = vars(species))+
  scale_color_brewer(palette = "Set1")+
  theme_bw()+
  ylab("Recovery Log Response Ratio")+
  xlab("Time (month)")+
  theme(strip.text.x = element_text(size = 10, face = "italic"),
        axis.title.y = element_text(size = 14, vjust = +1.5),
        axis.title.x = element_text(size = 14, vjust = -0.75),
        axis.text.y = element_text(size = 11),
        axis.text.x = element_text(size = 11)
      )

plot_grid(mo_rel, co_rel, ncol = 1, labels = c("A)", "B)"))


## save figure
pdf("../figures/all_relative_data.pdf", width = 10, height = 8)

plot_grid(mo_rel, co_rel, ncol = 1, labels = c("A)", "B)"), align = "h", axis = "l")

invisible(dev.off())

```

## Analysis

Explore how proportion seed removal is a function of seed size, trap method and time with a random effect for block nested within site.

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
## data cleanup

# add seed mass data into big dataset.
mass_all <- rbind(mo_mass_small, co_mass_small)
dat_full <- left_join(dat_all, mass_all)

# make month numeric and fix month timing
dat_full$time_month <- as.numeric(dat_full$time_month)
dat_full$time_month[dat_full$time_month == 1] <- 0.25
dat_full$time_month[dat_full$time_month == 2] <- 1
dat_full$time_month[dat_full$time_month == 3] <- 2


# remove the post-sticky treatment from MO
dat_full <- subset(dat_full, treatment != "post-sticky")

# calculate the number not recovered
dat_full$number_lost <- dat_full$number_original - dat_full$number_recovered

# treatment as factor
dat_full$treatment <- as.factor(dat_full$treatment)
```

```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
# Statistical Model
mod1 <- glmmTMB(pct_recovered ~ treatment * mean_1seed + time_month + (1|site/block), 
                data = dat_full, 
                weights = number_original, 
                family = binomial)
summary(mod1) 

```



Summary stats and main text figure to represent the significant trap type * seed mass interaction using model predicted values.

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

#Predict the data from the statistical model
out.pred <- predict(mod1, type = 'response')


predicted <- data.frame(pred = out.pred,
                        mass = dat_full$mean_1seed,
                        time = dat_full$time_month,
                        treatment = dat_full$treatment,
                        num_rec = dat_full$number_recovered,
                        num_lost = dat_full$number_lost)

## summary stats
predicted %>%
  group_by(treatment) %>%
  dplyr::summarize(mean_recovered = mean(pred),
                   sd_recovered = sd(pred))


palatte <- c("#CC79A7","#E69F00" )

predicted$treatment <- factor(predicted$treatment, levels = c("carpet", "sticky"), labels = c("astroturf", "sticky"))



#graph predicted data and binomial smooths
mass <- ggplot(predicted, aes(x = mass, y = pred, 
                     color = treatment))+
          geom_point(position = position_jitter(width = 0.0002), alpha = 0.5)+
          theme_bw()+
          scale_colour_manual(values = palatte)+
          labs(y= "Proportion of Seeds Recovered", x = "Seed mass", color = "Trap Type")+ 
          stat_smooth(method = "glm", 
                      method.args = list(family = "binomial"))+  
          theme(axis.title.y = element_text(size = 14, vjust = +1.5),
                axis.title.x = element_text(size = 14, vjust = -0.75),
                axis.text.y = element_text(size = 11),
                axis.text.x = element_text(size = 11)
      )

mass

## save figure
pdf("../figures/recovery_by_mass2.pdf", width = 6, height = 5)

mass

invisible(dev.off())

```

Supplemental figure to look at the predicted outcome for all factors (trap type, mass, time and site)

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# Predicted figure for all factors
pred <- ggpredict(mod1, terms=c("mean_1seed", "treatment", "site", "time_month"), type="random")
pred_all <- plot(pred)
pred_all


## save figure
pdf("../figures/all_predicted.pdf", width = 6, height = 5)

pred_all

invisible(dev.off())

```


## 1 month vs 1 month post sticky

```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
dat_test <- dat_all[,1:8]
dat_test <- subset(dat_test, site == "mo" & time_month == 1)

dat_test$pct_recovered <- dat_test$number_recovered/dat_test$number_original

ggplot(dat_test, aes(x = treatment, y = pct_recovered))+
  geom_boxplot()+
  theme_bw()


mod2 <- glmmTMB(pct_recovered ~ treatment + (1|block), 
                data = dat_test, 
                family = binomial)
summary(mod2) 
emmeans(mod2, pairwise~treatment)

```

